Graph Topology Learning from Binary Signals on Graph

### 2022-04-28 poster html copy/poster html.Rmd
minor changes on the poster

Data often has  an underlying structure or geometry that can be modeled as a signal on the vertices of a weighted, undirected graph. Traditionally, the emphasis was on using a underlying graph, or network, to understand the properties of signals over the vertices. Recently, there has been a surge in converse problem; learning a graph structure from a set of signals satisfying some constraints [@xia2021graph] [@dong2016learning] [ @ortega2018graph]. In previous research [@dong2016learning], [@tugnait2021sparse], [@pu2021learning], and [@saboksayr2021accelerated], signals on graph were assumed to follow multivariate Gaussian distributions, but there has been little exploration in learning a network from binomial signals. In this work, we develop a novel methodology that allows for learning graph topology given a set of binary signals on the graph.

```{r GTG, echo=FALSE,out.width="80%", out.height="20%",fig.cap="Circular Embedding of random Erd&#337;s-Reyni Graph with edge probability p = 0.2. Each edge weight follows uniform random distribution on (0,1). Grey nodes indicate a signal value of 0 and black nodes indicate a signal value of 1.",fig.show='hold',fig.align='center'}
knitr::include_graphics(c("pictures/gt_graph.png"))
```

@@

\begin{equation}
\label{eq:optimization}
\begin{aligned}
&\max_{\mathbf{A},h} \sum_{j = 1}^M\sum_{i=1}^N\left(y_{i,j}(\mathbf{A}[i,]h) - \log(1 + e^{\mathbf{A}[i,]h})\right) - \alpha \vert L \vert _ F \\
\textrm{s.t.} \quad  &\mathbf{A}_{i,j} = 0 \text{ if } i =j;\,\, \mathbf{A}_{i,j} \geq 0 \text{ if } i \neq j\\
                     &\max(h) \leq a ;  \,\, \min(h) \geq b \,\, \mathbb{1}h^T = 0  
\end{aligned}
\end{equation}
where $\alpha$ is a tuning parameter for controlling the sparsity and $\mathbf{A}$ and $b$ are tuning parameters for restricting the spread of the values of $h$. 
<hr />
\begin{align*}
&\textbf{Algorithm: Binary Signal Representation} \\
&1) \textbf{ Input: } \text{ Input a signal } \mathbf{Y}.\\
&2) \textbf{ Output: } \text{ Output an estimated } \mathbf{A}.\\
&3) \textbf{ Initialization: } h_{i,0} \sim \mathcal{N}(0,1) \text{ for } i = 1,\ldots, N \\
&4) \textbf{ for } t = 1, \ldots, iter:\\
&5) \textbf{ Update } \mathbf{A} \textbf{ given } h:\\
&6) \,\,\,\, \textit{ Fix h in Optimization Program and solve \mathbf{A}} \\
&7) \textbf{        Update } h \textbf{ given } \mathbf{A}:\\
&8) \,\,\,\, \textit{ Fix } \mathbf{A} \text{ in Optimization Program and solve h} \\
&9) \textbf{ end for }
\end{align*}
<hr />

# Results

In each iteration of step 5) and 6) of Algorithm 1, the optimization program gives a Disciplined Concave Program that can be solved efficiently in Python with the CVXPY library.

##### Synthetic Data Generation

Let $\mathbf{A}$ be the adjacency matrix of the graph in Figure \@ref(fig:GTG), and $h \in \mathbb{R}$ be such that $h_i \sim \mathcal{N}(0,1)$. Set $p = \mathbf{A}h$, and compare this to a vector $t$, where each $t_i$ follows a random uniform distribution $\mathcal{U}(0,1)$. If $t_k < p_k$, then node $k$ takes the signal value of $1$, and $0$ otherwise. We generate $M = 100$ synthetic signals, and use them to learn an adjacency matrix. The learned graph is found with parameter values $\alpha = .2$, $a = 1$ and $b = -1$.

@@

```{r HM, echo=FALSE,out.width="49%", out.height="20%",fig.cap="(Left) A heatmap of the ground truth adjacency matrix. (Right) A heatmap of the estimated adjacency matrix. We can see the learned adjacency matrix shares some similar features to the ground truth, but the learned adjacency matrix is sparser than the ground truth.",fig.show='hold',fig.align='center'}
knitr::include_graphics(c("pictures/gt_A_heatmap.png","pictures/est_A_heatmap.png"))
```

```{r EAG, echo=FALSE,out.width="80%", out.height="20%",fig.cap="Circular Embedding of Estimated graph. Note the overall weight of each edge is smaller than the edge weights of the ground truth.",fig.show='hold',fig.align='center'}
knitr::include_graphics(c("pictures/est_graph.png"))
``` 
# Discussion

To our knowledge, this is the first work on learning graph topology from binary signals. We have developed a method to effectively learn the graph adjacency matrix from binary signals. This model can be easily extended for learning binomial signals in general. Incorporating parameter optimization techniques would be a natural improvement on this model. Moreover, it would be desirable to develop scalable algorithms that can be used to learn a topology for a large graph from signals. Recently, in @saboksayr2021accelerated the author shows fast proximal-gradient iterations can be applied to the framework given by @kalofolias2016learn improving the over-all runtime.

```{r, include=FALSE}
knitr::write_bib(c('knitr','rmarkdown','posterdown','pagedown'), 'packages.bib')
```
# Acknowledgement
This work is supported by Research, Scholarship, and Creative Activity (RSCA) Award 2021-2022 through Chico State Enterprises. Thank you to Robin Donatello for approving this work.

>>>>

Data often has  an underlying structure or geometry that can be modeled as a signal on the vertices of a weighted, undirected graph. Traditionally, the emphasis was on using a underlying graph, or network, to understand the properties of signals over the vertices. Recently, there has been a surge in converse problem; learning a graph structure from a set of signals satisfying some constraints [@dong2016learning] [ @ortega2018graph]. In previous research [@dong2016learning], [@tugnait2021sparse], and [@saboksayr2021accelerated], signals on graph were assumed to follow multivariate Gaussian distributions, but there has been little exploration in learning a network from binomial signals. In this work, we develop a novel methodology that allows for learning graph topology given a set of binary signals on the graph as shown in Figure \@ref(fig:GTG).

```{r GTG, echo=FALSE,out.width="90%", out.height="25%",fig.cap="Circular Embedding of random Erd&#337;s-Reyni Graph with edge probability p = 0.2. Each edge weight follows uniform random distribution on (0,1). Grey nodes indicate a signal value of 0 and black nodes indicate a signal value of 1.",fig.show='hold',fig.align='center'}
knitr::include_graphics(c("pictures/gt_graph.png"))
```

@@

\begin{equation}
\label{eq:optimization}
\begin{aligned}
&\max_{\mathbf{A},h} \sum_{j = 1}^M\sum_{i=1}^N\left(y_{i,j}(\mathbf{A}[i,]h) - \log(1 + e^{\mathbf{A}[i,]h})\right) - \alpha \vert L \vert _ F \\
\textrm{s.t.} \quad  &\mathbf{A}_{i,j} = 0 \text{ if } i =j;\,\, \mathbf{A}_{i,j} \geq 0 \text{ if } i \neq j\\
                     &\max(h) \leq a ;  \,\, \min(h) \geq b \,\, ; \mathbb{1}h^T = 0  
\end{aligned}
\end{equation}
where $\alpha$ is a tuning parameter for controlling the sparsity and $\mathbf{A}$ and $b$ are tuning parameters for restricting the spread of the values of $h$.

We presented our optimization algorithm below. In each iteration of step 5) and 6) of Algorithm 1, the optimization program gives a Disciplined Concave Program that can be solved efficiently in Python with the CVXPY library.

<hr />
\begin{align*}
&\textbf{Algorithm 1: Binary Signal Graph Learning} \\
&1) \textbf{ Input: } \text{ Input a signal } \mathbf{Y}.\\
&2) \textbf{ Output: } \text{ Output an estimated } \mathbf{A}.\\
&3) \textbf{ Initialization: } h_{i,0} \sim \mathcal{N}(0,1) \text{ for } i = 1,\ldots, N \\
&4) \textbf{ for } t = 1, \ldots, iter:\\
&5) \textbf{ Update } \mathbf{A} \textbf{ given } h:\\
&6) \,\,\,\, \textit{ Fix }  h \textit{ in Optimization Program and solve } \mathbf{A} \\
&7) \textbf{        Update } h \textbf{ given } \mathbf{A}:\\
&8) \,\,\,\, \textit{ Fix } \mathbf{A} \text{ in Optimization Program and solve h} \\
&9) \textbf{ end for }
\end{align*}
<hr />

# Results
##### Synthetic Data Generation
Let $\mathbf{A}$ be the adjacency matrix of the graph in Figure \@ref(fig:GTG), and $h \in \mathbb{R}$ be such that $h_i \sim \mathcal{N}(0,1)$. Set $p$ as a logistic function of $\mathbf{A}h$, and compare $p$ to a vector $t$, where each $t_i$ follows a random uniform distribution $\mathcal{U}(0,1)$. If $t_k < p_k$, then node $k$ takes the signal value of $1$, and $0$ otherwise. We generate $M = 100$ synthetic signals, and use them to learn an adjacency matrix.

@@

The learned graph is found with parameter values $\alpha = .2$, $a = 1$ and $b = -1$ by implementing Algorithm 1. We can see the learned adjacency matrix shares some similar features to the ground truth (Figure \@ref(fig:HM)), but the learned adjacency matrix is sparser than the ground truth (Figure \@ref(fig:HM) and Figure \@ref(fig:EAG)).

```{r HM, echo=FALSE,out.width="49%", out.height="20%",fig.cap="(Left) A heatmap of the ground truth adjacency matrix. (Right) A heatmap of the estimated adjacency matrix. ",fig.show='hold',fig.align='center'}
knitr::include_graphics(c("pictures/gt_A_heatmap.png","pictures/est_A_heatmap.png"))
```

```{r EAG, echo=FALSE,out.width="60%", out.height="10%",fig.cap="Circular Embedding of Estimated graph. Note the overall weight of each edge is smaller than the edge weights of the ground truth.",fig.show='hold',fig.align='center'}
knitr::include_graphics(c("pictures/est_graph.png"))
``` 
# Discussion
To our knowledge, this is the first work on learning graph topology from binary signals. We have developed a method to effectively learn the graph adjacency matrix from binary signals. This model can be easily extended for learning binomial signals in general. Incorporating parameter optimization techniques would be a natural improvement on this model. Moreover, it would be desirable to develop scalable algorithms that can be used to learn a topology for a large graph from signals. Recently, in @saboksayr2021accelerated the author shows fast proximal-gradient iterations can be applied to the framework given by @kalofolias2016learn improving the over-all runtime.

```{r, include=FALSE}
knitr::write_bib(c('knitr','rmarkdown','posterdown','pagedown'), 'packages.bib')
```
# Acknowledgement
This work is supported by Research, Scholarship, and Creative Activity (RSCA) Award 2021-2022 through Chico State Enterprises. Thank you to Robin Donatello for approving this work for Data Science Capstone Project.
